---
title: 'P8180 - SQL Assignment #9'
author: "Jon Brock"
date: "2/20/2020"
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tibble.width = Inf)
```

```{r load packages, message = FALSE}
library(tidyverse)
library(sqldf)
```

```{r data import}
demo <- read.csv("./NHANES_Demographics.csv") %>% janitor::clean_names()
tri <- read.csv("./NHANES_Triglycerides.csv") %>% janitor::clean_names()
```

```{r df to tibble}
demo_df <- as_tibble(demo)
tri_df <- as_tibble(tri)
```

```{r data cleaning}
demo_clean <- mutate(demo_df,
                seqn = respondent_sequence_number,
                age = age_in_years_at_screening,
                race = as_factor(race_hispanic_origin_w_nh_asian),
                pregnant = as_factor(pregnancy_status_at_exam),
                household_income = annual_household_income,
                family_income = annual_family_income,
                gender = as_factor(gender)) %>% 
            select(seqn, age, race, pregnant, household_income, family_income, gender)

tri_clean <- mutate(tri_df,
                seqn = respondent_sequence_number,
                ldl_mmol = ldl_cholesterol_mmol_l,
                ldl_mgdl = ldl_cholesterol_mg_d_l,
                tri_mmol = triglyceride_mmol_l,
                tri_mgdl = triglyceride_mg_d_l) %>% 
            select(seqn, ldl_mmol, ldl_mgdl, tri_mmol, tri_mgdl)

head(demo_clean)
head(tri_clean)
```

1. Write a query that would allow you to fill out table 1 and assign the results to an object called table1.

```{r problem #1}
table1 <- sqldf(
            "SELECT race, COUNT(race) AS freq, ROUND(AVG(age), 1) AS mean_age
             FROM demo_clean
             GROUP BY race")
table1
```

2. Show the distribution of race by gender and display the race/gender combinations from highest to lowest frequency.  Note: when using SQL in R, you *can* refer to column aliases outside of the SELECT clause.

```{r problem #2}
table2 <- sqldf(
            "SELECT gender, race, COUNT(race) AS race_counts
             FROM demo_clean
             GROUP BY gender, race
             ORDER BY race_counts DESC")
table2
```

3. Count the number of women who were pregnant at the time of screening. Use the column alias preg_at_screen.

```{r problem #3}
table3 <- sqldf(
            "SELECT COUNT(pregnant) AS preg_at_screen
             FROM demo_clean
             GROUP BY gender
             HAVING gender = 1")
table3
```

4. How many men refused to provide annual household income?

```{r problem #4}
table4 <- sqldf(
            "SELECT COUNT(gender) AS male_count
             FROM demo_clean
             GROUP BY household_income
             HAVING household_income = 77")
table4
```

5. What is the mean LDL level (mg/dL) for men and women? Use column alias mean_ldl and round results to 1 decimal place.

```{r problem #5}
table5 <- sqldf(
            "SELECT gender, ROUND(AVG(ldl_mgdl), 1) AS mean_ldl
             FROM demo_clean AS d
             LEFT JOIN tri_clean AS t
             ON d.seqn = t.seqn
             GROUP BY gender")
table5
```

6. Display the minimum and maximum triglyceride levels (mmol/L) for each race. Use column aliases min_tri and max_tri.

```{r problem #6}
table6 <- sqldf(
            "SELECT race, MIN(tri_mmol) AS min_tri, MAX(tri_mmol) AS max_tri
             FROM demo_clean AS d
             LEFT JOIN tri_clean AS t
             ON d.seqn = t.seqn
             GROUP BY race
             HAVING tri_mmol IS NOT NULL OR tri_mgdl IS NOT NULL")
table6
```

7. Create a new data frame that can be used for future analyses that combines all demographic data and any matching triglyceride data. Call it demo_tri.

```{r problem #7}
demo_tri <- sqldf(
        "SELECT d.*, t.tri_mmol, t.tri_mgdl
         FROM demo_clean AS d
         LEFT JOIN tri_clean AS t
         ON d.seqn = t.seqn
         WHERE tri_mmol IS NOT NULL OR tri_mgdl IS NOT NULL") %>% 
    as_tibble()
demo_tri
```